\Title

### 2013-02-03 synmut.tex
text small changes

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the immune system. Synonymous mutations do not have an
immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intra-patient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles rarely reach
fixation.
Simulations suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
detect a negative correlation between the fixation of an allele and its
involvement in RNA stem-loop structures.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines). Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
however, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011}.
This is expected to be particularly important in compact viruses like HIV. 
In addition to maintaining protein function and avoiding the adaptive immune
recognition, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but while positive
selection through the immune system be mainly restricted to changes in the
amino acid sequence.

%SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE
%FOR SUPPLEMENT? YES

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We then
compare our observations to computational models of HIV evolution and derive
estimates for the effect synonymous mutations have on fitness.
Hitchhiking of synonymous mutations with escape mutations is due to the small
recombination rate of HIV~\citep{neher_recombination_2010,
batorsky_estimate_2011}. Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft}A\&B show 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5 using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

@@

When interpreting these results for the fixation probabilities, it is important
to distinguish random mutations, and polymorphisms at a given frequency that has
already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by a chance event (genetic drift or hitch-hiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations are neutral
-- only those mutations observed at high frequencies tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. Panel A shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles are lost (blue
line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. As a control, the
average over non-observed but potentially available polymorphisms lies between
the two curves (green line), as expected (because only some of these mutations
will interfere with stem loop formation).
To test the hypothesis that mutations in C2-V5 are lost since break stems in the
conserved between the variable loops, we split the synonymous mutations in the
extended V1-V5 region further into conserved and variable regions and found that
the biggest depression in fixation probability is observed in the conserved
stems, while the variable loops show little deviation from the neutral
signature, see \FIG{SHAPEB}. This is consistent with important stem structures
in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\subsection{Deleterious mutations are brought to high frequency by hitch-hiking}

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitch-hiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the deleterious
effect of synonymous mutations, the fraction of synonymous mutations that are
deleterious, the escape rate of adaptive mutations, and the frequency of
escapes.  Among all simulations, we select the ones that show $A_\text{syn}$ and
$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
probability of synonymous mutations but, simulteneously, a moderately high
synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$
(twofold variation around the observed value). This selection imposes a strong
constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
The results are plausible:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), a depression in $\pfix$ will be
visible only if deleterious mutations are pervasive; (b) in order to
hitchhike, the deleterious effect size has to be less than the escape rate, otherwise the
double mutant has little or no fitness advantage; (c) mutations with a 
deleterious effect weaker than approximately $0.001$ behave neutrally
consistent with the typical coalescent times observed in HIV.

@@

The former hypothesis can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative hypothesis, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc.~is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements \comment{Supplementary}.
In real HIV infections, both mechanisms are likely to be playing a role.

@@

Contrary to na\"ive expectations, the adaptive escape mutations don't seem to be
unconditionally beneficial. Otherwise we would observe almost sure fixation of
nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation. 
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield
the neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino-acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the nonsynonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to (a) time dependent environment
through an immune system that is catching up, or (b) competition between mutations
that mediate escape within the same epitope. We explored both of these
possibilities and find that both produce the desired effect. Furthermore, there
is evidence in support of both of these hypothesis. Serum from HIV infected
individuals tyically neutralizes the virus that dominated the population a few
(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
cease to be beneficial after a few months and might revert if they come with a
fitness cost. Deep sequencing of regions of \env{} after antibody escape have
revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them don't provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point mutations rapidly. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.

>>>>

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not have an
adaptive immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intrapatient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles are rarely
fixed.
Computer models suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the V loops are more likely to be lost than other synonymous changes, hinting at
a direct fitness effect of such stem-loop structures in the HIV RNA.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes.
On the one hand, these processes often involve some kind of RNA secondary structure.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
On the other hand, nucleotide-level fitness effects have been observed beyond RNA structure as
well.
In most species, it is  known that synonymous codons don't evolve completely neutrally
but that some codons are favored over others \citep{plotkin_synonymous_2011};
this is expected to be particularly important in compact viruses like HIV. 
Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}; nonetheless, positive
selection via the host adaptive immune system is restricted to changes in the
amino acid sequence.

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious.
We argue that synonymous mutations reach high frequencies via genetic hitchhiking
thanks to the small recombination rate of HIV~\citep{neher_recombination_2010,
batorsky_estimate_2011}.
We then
compare our observations to computational models of HIV evolution and derive
estimates for the effect synonymous mutations have on fitness.
Extending the analysis of fixation probabilities to the
nonsynonymous mutations, we show that time dependent selection or strong
competition of escape mutations inside the same epitope are necessary to explain
the observed patterns of fixation and loss.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
less the neutral expectation (blue line) -- a point to which we will come back below.

@@

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms at a given frequency,
which have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by chance (genetic drift or hitchhiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations are neutral
-- only those mutations observed at high frequencies, which have been
experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red line) are systematically larger than of alleles are lost (blue
line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. As a control, the
average over non-observed but potentially available polymorphisms lies between
the two curves (green line), as expected (because only some of these mutations
would interfere with stem-loop formation).
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we split the synonymous mutations in the
extended V1-V5 region further into conserved and variable regions and found that
the biggest depression in fixation probability is observed in the conserved
stems, while the variable loops show little deviation from the neutral
signature, see \FIG{SHAPEB}. This is consistent with important stem structures
in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

\subsection{Deleterious mutations are brought to high frequency by hitchhiking}

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the deleterious
effect of synonymous mutations, the fraction of synonymous mutations that are
deleterious, the fitness effect of adaptive mutations (escape rate) and the frequency of
escapes. Among all simulations, we select the ones that show $A_\text{syn}$ and
$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
probability of synonymous mutations but, simulteneously, a moderately high
synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
0.010$. This selection imposes a strong
constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
with effect size $|s_d| \sim 0.002$, is required. This result fits well the
expectation based on the fixation/extinction times above (see \FIG{fixp1}).
The results are plausible:
(a) since high frequency bins are enriched in neutral mutations (it is easier
for them to hitchhike), a depression in $\pfix$ will be
visible only if deleterious mutations are pervasive; (b) in order to
hitchhike, the deleterious effect size has to be much smaller than the escape
rate, otherwise the
double mutant has little or no fitness advantage; (c) mutations with a 
deleterious effect weaker than approximately $0.001$ behave neutrally
consistent with the typical coalescent times observed in HIV.

@@

The former hypothesis can be formulated as follows: if the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average time needed by an escape mutant to cross frequencies of order one.
In the alternative hypothesis, several different escape
mutations within the same epitope might arise almost simultaneously and start to
spread. Their benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency rapidly, while the one with the smallest cost in terms of replication,
packaging, etc.~is most likely to eventually fix. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
We tested both hypotheses in our simulation framework and they both seem to
achieve a reduced, neutral-like fixation probability while maintaining a high
rate of substitutions, compatibly with HIV measurements (see supplementary
materials).
In real HIV infections, both mechanisms are likely to be playing a role, as they
are not mutually exclusive.

@@

Contrary to na\"ive expectations, the adaptive escape mutations do not seem to be
unconditionally beneficial. Otherwise we would observe almost sure fixation of
nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation. 
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield
the neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino-acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the nonsynonymous mutations is much more deleterious (see supplementary
materials).

Alternatively, the lack of fixation could be due to (a) time dependent environment
through an immune system that is catching up, or (b) competition between mutations
that mediate escape within the same epitope. We explore both of these
possibilities and find that both produce the desired effect. Furthermore, there
is experimental evidence in support of both of these hypotheses. Serum from HIV infected
individuals tyically neutralizes the virus that dominated the population a few
(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
cease to be beneficial after a few months and might revert if they come with a
fitness cost. Deep sequencing of regions of \env{} after antibody escape have
revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them don't provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point mutations rapidly. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.
